Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2022-01-03 analysis/paper/paper.Rmd
complete draft

Predation of emigrating juvenile salmonids by piscivorous fishes is perhaps an under-appreciated source of mortality hindering population recovery in the Columbia River basin. Reservoirs and tailraces associated with dams on the Snake and Columbia rivers have created favorable conditions for fishes known to consume juvenile salmonids, including the Northern Pikeminnow *Ptychocheilus oregonsis*. Indeed, Northern Pikeminnow have been estimated to consume 8% of the approximately 200 million juvenile salmonids emigration through those rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine predation by Northern Pikeminnow on juvenile salmonids, focusing on Chinook Salmon *Oncorhynchus tshawytscha*, in a slow-water reach of the Salmon River, Idaho known as Deadwater Slough. The reach contains historically important overwinter rearing and spring emigration habitat for Chinook Salmon and other ESA-listed species, and was formed by the failure of a small mining dam in a reservoir in the late 19th century. We estimate the abundance of Northern Pikeminnow in Deadwater Slough during peak emigrations of Chinook salmon, and further, use a bioenergetics approach to estimate the consumption potential of the Northern Pikeminnow population to assess potential impacts to Chinook Salmon emigrants. Finally, we evaulate potential impacts to adult returns of local populations. ***We found that... (insert concluding findings here.*** It is likely that predation of juvenile Chinook Salmon in Deadwater Slough has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River MPG hindering local recovery efforts.

@@

The Snake and Columbia rivers drain a network of inland tributaries that provide essential spawning and rearing habitat for anadromous Pacific salmon *Oncorhynchus* spp. and steelhead *O. mykiss*. These fishes were historically abundant throughout the drainage basin and consequently have tremendous ecological, cultural, and economic value. Over the last century, salmonid stocks throughout the Snake and Columbia rivers have undergone significant declines related to habitat modification in the freshwater environment (e.g., removal of beavers, river channel simplification, water withdrawals, hydrosystem development) and poor marine conditions, both affecting their survival and recruitment at multiple life stages. Consequently, action agencies have made considerable investments in the rehabilitation of tributary ecosystems and evaulation of factors attributed to the species decline [@Roni2018].

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` kilometer long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is located in a river reach containing historically important overwinter rearing and spring emigration habitat for juvenile Chinook Salmon *O. tshawytscha* and is inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*. Further, the slough is part of the migratory pathway of the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead.

Quantifying mortality associated with piscine predation in upstream habitats of the Salmon River is of particular interest for spring/summer Chinook Salmon in the Upper Salmon major population group (MPG). The Salmon River was historically the most productive tributary for spring/summer Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Currently, the Upper Salmon MPG supports eight independent, extant spring/summer Chinook Salmon populations, including Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. The Upper Salmon MPG are stream-type Chinook Salmon and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency. Because the Deadwater Slough supports rearing for DSR migrants and is part of the migration corridor for NRR migrants, and juveniles from all Upper Salmon populations must rear in or migrate through it, it represents a significant habitat for the Upper Salmon MPG.

Recent studies have attempted to examine the downstream movement, distribution, and (apparent) survival of juvenile salmonid emigrants through the Salmon River, including the Deadwater Slough reach. Sockeye Salmon migrating through the mainstem Salmon River during spring were estimated to have substantially reduced survival in the Deadwater Slough reach compared to adjacent reaches [@Axel2015]. Similarly, emigrating DSR Chinook Salmon had a 10% reduction in transition probability through the Deadwater Slough compared to surrounding reaches during the fall and early winter months [@Ackerman2018; @Porter2019]. Low survival in these cases was attributed to low-velocity water delaying movement rates, thereby increasing predation risk.

@@

1. Estimate the abundance (or relative abundance) of potential predators in the Deadwater Slough during the peaks of fall (DSR) and spring (NRR) juvenile emigrations.
2. Document predation on juvenile Chinook Salmon during the emigration periods using gastric lavage.
3. Use a bioenergetics approach to estimate consumption potential (grams) of the Northern Pikeminnow population at Deadwater Slough to assess their potential impacts to DSR and NRR Chinook Salmon emigrants.

We also consider how consumption of juvenile Chinook Salmo emigrants at Deadwater Slough by Northern Pikeminnow may impact adult returns to the Upper Salmon MPG.

@@

The Deadwater Slough is an approximately 1.5 km section of the mainstem Salmon River located approximately 6 river km downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event that deposited substantial amounts of sediment at the confluence of the Salmon River [@Emerson1973], thereby creating a slow, deep section in the river, spanning approximately `r round(ddw_acres)` acres (`r round(ddw_hectares)` hectares), and averaging `r round(ddw_avg_width_m)` m width.

## Northern Pikeminnow Abundance

@@

Multiple capture methods were employed during the initial fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. Ineffective methods were then abandoned in subsequent years. <!-- For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. -->After evaluating all methods, angling proved to be the most effective method for capturing piscine predators while also minimizing potential impacts to ESA-listed adult steelhead present in the Salmon River during our sampling efforts.<!--, but recognizing that angling can have both species and size selectivity (bias).--> Therefore, the following analyses will focus on fish captured by angling, unless otherwise noted. During the fall 2019 effort it additionally became clear that Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and so is the focus of our study.

@@

where $M$, $n$, and $m$ are indexed by each sample, $i$. Here, sample $i$ is defined as each day. The Schnabel estimator does not have an associated standard error; however, the 95% confidence intervals were calculated. Lastly, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

@@

Northern Pikeminnow in Deadwater Slough were not only abundant, but anglers reported that many were large-bodied. For that reason, we calculated the proportional stock density (PSD) for Northern Pikeminnow in Deadwater Slough, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality_sized" (larger) individuals and "stock-sized" individuals.

@@

where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$. Stock and quality size definitions vary by species. We calculated the PSD for Northern Pikeminnow in Deadwater Slough using `r prettyNum(s_size, digits = 0)` mm TL for stock-size and `r prettyNum(q_size, digits = 0)` mm TL for quality-size [@Winther2020].

## Gastric Lavage

@@

## Bioenergetics

@@

The largest uncertainty among model parameters was the proportion of the Northern Pikeminnow diet consisting of fish versus non-fish prey items (e.g., invertebrates). Therefore, we conducted nine model runs with varying proportions of fish in the diet ranging from 10-90%, in 10% increments. Energy densities for invertebrates was fixed at 3,000 J/g. Because we were unable to differentiate juvenile Chinook Salmon from other fish prey consumed, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. Separate models were also run for a 77-day fall period from September 15 - November 30 and a 92-day period in the spring from March 1 - May 31 to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG. This resulted in 18 separate model runs (9 diet scenarios and 2 seasons). Each model run provided an estimate of the amount (grams) of fish consumed by an individual Northern Pikeminnow during the fall or spring time periods.

To estimate the total biomass of Chinook Salmon potentially consumed by all Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated predator population size during each period (fall and spring). The estimate of total biomass was converted to the total number of fish consumed each period by dividing by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The average weights of DSR and NRR emigrants were taken from fish caught at seven rotary screw traps located throughout the Upper Salmon MPG upstream of Deadwater Slough during the fall and spring time periods. This calculation assumes that all fish prey items consumed by Northern Pikeminnow are, on average, the same size as DSR and NRR emigrants.

@@

## Northern Pikeminnow Abundance

@@

Mark-recapture abundance estimates for Northern Pikeminnow in Deadwater Slough during fall 2019 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` and for fall 2020 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` (Table \@ref(tab:abund-tab)). In both cases, the multiple census estimators were larger than the single census estimators. Confidence intervals for the fall 2020 effort were more varied, likely resulting from fewer recaptures in that year.

@@

## Gastric Lavage

We examined the stomach contents of Northern Pikeminnow from Deadwater Slough using gastric lavage. To validate the efficacy of the method, we euthanized 75 Northern Pikeminnow after gastric lavage was completed and removed the remaining food items via dissection. We found that gastric lavage successfully removed food items in all dissected individual, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in some form in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) of those. Northern Pikeminnow captured during the spring 2021 effort had a higher proportion of diets with fish contents (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Spring_2021"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Spring_2021"], 1)`%) than fish captured in the fall 2019 (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Fall_2019"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Fall_2019"], 1)`%) and fall 2020 (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Fall_2020"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Fall_2020"], 1)`%) (Table \@ref(tab:gastric-tab)). For those individuals with stomach contents, the average wet weight of total contents was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; range = `r round(min(stmch_df$stomach_contents_weight), 2)`-`r round(max(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants were `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall stomach contents examined.

## Bioenergetics

We estimated an individual, average-size (`r avg_fall_NPM` mm) Northern Pikeminnow to consume `r median(fall_g_summ$cumm_g)` g (range `r min(fall_g_summ$cumm_g)` - `r max(fall_g_summ$cumm_g)` g) of fish during the DSR emigration to maintain their body size, depending on the proportion of diet consisting of fish versus invertebrates (Figure \@ref(fig:bio-p)). Similarly, we estimated an average-size (`r avg_spring_NPM` mm) Northern Pikeminnow to consume `r median(spring_g_summ$cumm_g)` g (range `r min(spring_g_summ$cumm_g)` - `r max(spring_g_summ$cumm_g)` g) of fish during the NRR emigration (Figure \@ref(fig:bio-p)). Given an average weight of `r avg_DSR_g` g for DSR emigrants, that would equate to `r round(median(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)` (range `r round(min(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)` - `r round(max(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)`) fish consumed by an individual during fall and an additional `r round(median(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)` (range `r round(min(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)` - `r round(max(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)`) fish consumed using an average weight of `r avg_NRR_g` g for NRR emigrants.

@@

Across both seasons and all diet scenarios, we estimated the total number of fish consumed to be between `r prettyNum(round(min(total_impacts$smolt_N), 0), big.mark = ",")` and `r prettyNum(round(max(total_impacts$smolt_N), 0), big.mark = ",")` (Table \@ref(tab:adult-imp)). The estimated number of fish consumed by a Northern Pikeminnow begins to taper off between the 30-90% diet scenarios. So although we are uncertain of the diet proportions consisting of fish versus invertebrates, the impact of that uncertainty is somewhat constrained.
<!--
## Juvenile Chinook Salmon Abundance
For context, throughout the migration periods of DSR and NRR life histories it is estimated that 113,976 individuals are produced and therefore available within the upper Salmon River basin (Table x).

@@

## Gastric Lavage

>>>>

Predation of emigrating juvenile salmonids by piscivorous fishes is perhaps an under-appreciated source of mortality hindering population recovery in the Columbia River basin. Reservoirs and tailraces associated with dams on the Snake and Columbia rivers have created favorable conditions for fishes known to consume juvenile salmonids, including the Northern Pikeminnow *Ptychocheilus oregonsis*. Indeed, Northern Pikeminnow have been estimated to consume 8% of the approximately 200 million juvenile salmonids emigration through those rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above these reservoir systems. In this study, we examine predation by Northern Pikeminnow on juvenile salmonids, focusing on Chinook Salmon *Oncorhynchus tshawytscha*, in a slow-water reach of the Salmon River, Idaho known as Deadwater Slough. The reach contains historically important overwinter rearing and spring emigration habitat for Chinook Salmon and other ESA-listed species, and was formed by the failure of a small mining dam in a reservoir in the late 19th century. We estimate the abundance of Northern Pikeminnow in Deadwater Slough during peak emigrations of Chinook salmon, and further, use a bioenergetics approach to estimate the consumption potential of the Northern Pikeminnow population to assess potential impacts to Chinook Salmon emigrants. Finally, we evaulate potential impacts to adult returns of local populations. ***We found that... (insert concluding findings here.*** It is likely that predation of juvenile Chinook Salmon in Deadwater Slough has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River MPG hindering local recovery efforts.

@@

The Snake and Columbia rivers drain a network of inland tributaries that provide essential spawning and rearing habitat for anadromous Pacific salmon *Oncorhynchus* spp. and steelhead *O. mykiss*. These fishes were historically abundant throughout the basin and consequently have tremendous ecological, cultural, and economic value. Over the last century, salmonid stocks throughout the Snake and Columbia rivers have undergone significant declines related to habitat modification in the freshwater environment (e.g., removal of beavers, river channel simplification, water withdrawals, hydrosystem development) and poor marine conditions, both affecting their survival and recruitment at multiple life stages. Consequently, action agencies have made considerable investments in the rehabilitation of tributary ecosystems and evaulation of factors attributed to the species decline [@Roni2018].

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is located in a river reach containing historically important overwinter rearing and spring emigration habitat for juvenile Chinook Salmon *O. tshawytscha* and is inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*. Further, the slough is part of the migratory pathway of the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead.

Quantifying mortality associated with piscine predation in upstream habitats of the Salmon River is of particular interest for spring/summer-run Chinook Salmon (hereafter Chinook Salmon) in the Upper Salmon major population group (MPG). The Salmon River was historically the most productive tributary for Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Currently, the Upper Salmon MPG supports eight independent, extant populations, including Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. The Upper Salmon MPG are stream-type Chinook Salmon and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave their natal area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency. Because the Deadwater Slough supports rearing for DSR migrants and is part of the migration corridor for NRR migrants, and juveniles from all Upper Salmon populations must rear in or migrate through it, it represents a significant habitat for the Upper Salmon MPG.

Recent studies have examined the downstream movement, distribution, and (apparent) survival of juvenile salmonid emigrants through the Salmon River, including the Deadwater Slough reach. Sockeye Salmon migrating through the mainstem Salmon River during spring were estimated to have substantially reduced survival in the Deadwater Slough reach compared to adjacent reaches [@Axel2015]. Similarly, emigrating DSR Chinook Salmon had a 10% reduction in transition probability through the Deadwater Slough compared to surrounding reaches during the fall and early winter months [@Ackerman2018; @Porter2019]. Low survival in these cases was attributed to low-velocity water delaying movement rates, thereby increasing predation risk.

@@

1. Estimate the abundance (or relative abundance) of potential predators in the Deadwater Slough during the peaks of fall (DSR) and spring (NRR) juvenile emigrations.
2. Document predation on juvenile Chinook Salmon during the DSR and NRR emigration periods using gastric lavage.
3. Use a bioenergetics approach to estimate consumption potential (grams) of the Northern Pikeminnow population at Deadwater Slough to assess their potential impacts to DSR and NRR Chinook Salmon emigrants.

We also consider how consumption of juvenile Chinook Salmon emigrants at Deadwater Slough by Northern Pikeminnow may impact adult returns to the Upper Salmon MPG.

@@

The Deadwater Slough is an approximately `r round(ddw_rkms, 1)` km section of the mainstem Salmon River located approximately 6 river km downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event that deposited substantial amounts of sediment at the confluence of the Salmon River [@Emerson1973] forming an alluvial fan. The fan created a slow, deep section in the river, spanning approximately `r round(ddw_acres)` acres (`r round(ddw_hectares)` hectares), averaging `r round(ddw_avg_width_m)` m width, and resembling a small, narrow pond or reservoir.

## Northern Pikeminnow Population Size

@@

Multiple capture methods were employed during the initial fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. Ineffective methods were then abandoned in subsequent years. <!-- For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. -->After evaluating all methods, angling proved to be the most effective method for capturing piscine predators while also minimizing potential impacts to ESA-listed adult steelhead present in the Salmon River during our sampling efforts.<!--, but recognizing that angling can have both species and size selectivity (bias).--> Therefore, the following analyses will focus on fish captured by angling, unless otherwise noted. During the fall 2019 effort it additionally became clear that Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and so is our focus.

@@

where $M$, $n$, and $m$ are indexed by each sample, $i$. Here, sample $i$ is defined as each day. The Schnabel estimator does not have an associated standard error; however, the 95% confidence intervals were calculated. The Schnabel estimator is essentially a weighted average of a series of Lincoln-Peterson estimators (with a Champman modification). Lastly, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

@@

Northern Pikeminnow in Deadwater Slough were not only abundant, but anglers reported that many were large-bodied. For that reason, we calculated the proportional stock density (PSD) for Northern Pikeminnow in Deadwater Slough, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality-sized" (larger) individuals and "stock-sized" individuals:

@@

where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$. Quality and stock size definitions vary by species. We calculated the PSD for Northern Pikeminnow in Deadwater Slough using `r prettyNum(q_size, digits = 0)` mm TL for quality-size and `r prettyNum(s_size, digits = 0)` mm TL for stock-size [@Winther2020].

## Stomach Contents

@@

## Fish Consumption Potential

@@

The largest uncertainty among model parameters was the proportion of the Northern Pikeminnow diet consisting of fish versus non-fish prey items (e.g., invertebrates). Therefore, we conducted nine model runs with varying proportions of fish in the diet ranging from 10-90%, in 10% increments. Energy densities for invertebrates was fixed at 3,000 J/g. Because we were unable to differentiate juvenile Chinook Salmon from other fish prey consumed during model runs, we assume all fish prey have the same energy densities as juvenile Chinook Salmon i.e., juvenile Chinook Salmon were used as surrogates for all fish prey items. Separate models were also run for a 77-day fall period from September 15 - November 30 and a 92-day period in the spring from March 1 - May 31 to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG. This resulted in 18 separate model runs (9 diet scenarios and 2 seasons). Each model run provided an estimate of the total amount (grams) of fish consumed by an individual Northern Pikeminnow during the fall or spring time periods. To estimate the total biomass of fish potentially consumed by all Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated predator population size during each period (fall and spring).

## Potential Impacts to Chinook Salmon Populations

Multiplying the grams of fish consumed by an individual Northern Pikeminnow by the Northern Pikeminnow population size provides an estimate of total fish consumption potential for the population. However, the bioenergetics model runs do not differentiate juvenile Chinook Salmon from other fish prey items. Therefore, to understand potential impacts to local Chinook Salmon populations, we need to parse out the portion of total fish consumed that is juvenile Chinook Salmon. Unfortunately, estimates of the proportion of the Northern Pikeminnow fish prey that are juvenile Chinook Salmon are sparse, as well as information on the prevalence of juvenile Chinook Salmon relative to other fish prey items. Therefore, we instead consider hypothetical scenarios where Chinook Salmon are 1%, 10%, 30%, and 50% of fish prey consumed, by weight, which provide estimates of the total biomass (grams) of juvenile Chinook Salmon consumed by Northern Pikeminnow during the DSR and NRR emigration periods.

The estimate of total juvenile Chinook Salmon biomass consumed was converted to the total number of fish consumed each period by dividing by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The average weights of DSR and NRR emigrants were taken from fish caught at seven rotary screw traps located throughout the Upper Salmon MPG upstream of Deadwater Slough during the fall and spring time periods.

@@

## Northern Pikeminnow Population Size

@@

Mark-recapture abundance estimates for Northern Pikeminnow in Deadwater Slough during fall 2019 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` and for fall 2020 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` (Table \@ref(tab:abund-tab)). In both cases, the multiple census estimators were larger than the single census estimators. Confidence intervals for the fall 2020 effort were more varied, likely due to fewer recaptures during that year.

@@

## Stomach Contents

We examined the stomach contents of Northern Pikeminnow from Deadwater Slough using gastric lavage. To validate the efficacy of the method, we euthanized 75 Northern Pikeminnow after gastric lavage was completed and removed the remaining food items via dissection. We found that gastric lavage successfully removed food items in all dissected individuals, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in some form in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) of those. Northern Pikeminnow captured during the spring 2021 effort had a higher proportion of diets with fish contents (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Spring_2021"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Spring_2021"], 1)`%) than fish captured in the fall 2019 (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Fall_2019"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Fall_2019"], 1)`%) and fall 2020 (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Fall_2020"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Fall_2020"], 1)`%) (Table \@ref(tab:gastric-tab)). For those individuals with stomach contents, the average wet weight of total contents was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; range = `r round(min(stmch_df$stomach_contents_weight), 2)`-`r round(max(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants were `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall stomach contents examined.

## Fish Consumption Potential

We estimated an individual, average-size (`r avg_fall_NPM` mm) Northern Pikeminnow to consume `r median(fall_g_summ$cumm_g)` g (range: `r min(fall_g_summ$cumm_g)` - `r max(fall_g_summ$cumm_g)` g) of fish during the DSR emigration to maintain their body size, depending on the proportion of diet consisting of fish versus invertebrates (Figure \@ref(fig:bio-p)). Similarly, we estimated an average-size (`r avg_spring_NPM` mm) Northern Pikeminnow to consume `r median(spring_g_summ$cumm_g)` g (range: `r min(spring_g_summ$cumm_g)` - `r max(spring_g_summ$cumm_g)` g) of fish during the NRR emigration (Figure \@ref(fig:bio-p)). Given a mean Northern Pikeminnow population size of `r prettyNum(unique(adult_impact$N[adult_impact$season == "Fall"]), big.mark = ",")` from fall estimates and `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]), 0), big.mark = ",")` for spring 2021 and using the median value from diet scenarios (i.e., 50% of their diet consists of fish), we'd estimate that the Northern Pikeminnow population can consume `r prettyNum(median(fall_g_summ$cumm_g) * unique(adult_impact$N[adult_impact$season == "Fall"]), big.mark = ",")` g (range: `r prettyNum(min(fall_g_summ$cumm_g) * unique(adult_impact$N[adult_impact$season == "Fall"]), big.mark = ",")` - `r prettyNum(max(fall_g_summ$cumm_g) * unique(adult_impact$N[adult_impact$season == "Fall"]), big.mark = ",")` g) fish during the DSR emigration and an additional `r prettyNum(round(median(spring_g_summ$cumm_g) * unique(adult_impact$N[adult_impact$season == "Spring"]), 0), big.mark = ",")` g (range: `r prettyNum(round(min(spring_g_summ$cumm_g) * unique(adult_impact$N[adult_impact$season == "Spring"]), 0), big.mark = ",")` - `r prettyNum(round(max(spring_g_summ$cumm_g) * unique(adult_impact$N[adult_impact$season == "Spring"]), 0), big.mark = ",")` g) during the NRR emigration in Deadwater Slough. In total, the Northern Pikeminnow population can consume `r prettyNum(median(fall_g_summ$cumm_g) * unique(adult_impact$N[adult_impact$season == "Fall"]) + median(spring_g_summ$cumm_g) * unique(adult_impact$N[adult_impact$season == "Spring"]), big.mark = ",")` g of fish during the DSR and NRR emigrations to maintain their population size if we assume 50% of their diet consists of fish.

## Potential Impacts to Chinook Salmon Populations

1. Start with total g of fish consumed by Northern Pikemnnow population, summing DSR and NRR time periods.
2. Parse out portion of fish total g consumed that is juvenile Chinook Salmon (1%, 10%, 30%, 50%).
3. Divide by average weight of juvenile Chinook Salmon (10.6 g), the average of DSR (10.3 g) and NRR (10.9 g) emigrants.
4. Estimate of total juvenile Chinook Salmon consumed.
5. Multiply by SAR to estimate "adult equivalents".

<!--
Given an average weight of `r avg_DSR_g` g for DSR emigrants, that would equate to `r round(median(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)` (range: `r round(min(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)` - `r round(max(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)`) fish consumed by an individual during fall and an additional `r round(median(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)` (range: `r round(min(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)` - `r round(max(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)`) fish consumed using an average weight of `r avg_NRR_g` g for NRR emigrants.

@@

Across both seasons and all diet scenarios, we estimated the total number of fish consumed to be between `r prettyNum(round(min(total_impacts$smolt_N), 0), big.mark = ",")` and `r prettyNum(round(max(total_impacts$smolt_N), 0), big.mark = ",")` (Table \@ref(tab:adult-imp)). The estimated number of fish consumed by a Northern Pikeminnow begins to taper off between the 30-90% diet scenarios. So although we are uncertain of the diet proportions consisting of fish versus invertebrates, the impact of that uncertainty is somewhat constrained.
-->

<!--
## Juvenile Chinook Salmon Abundance
For context, throughout the migration periods of DSR and NRR life histories it is estimated that 113,976 individuals are produced and therefore available within the upper Salmon River basin (Table x).

@@

## Stomach Contents
